A comparison of design-based and model-based approaches for spatial data.

### 2021-10-10 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.Rmd
revised text for first conceptual figure.

Treating the data as fixed and incorporating randomness through the sampling design yields estimators having very few other assumptions. Confidence intervals for these types of estimators are typically derived using limiting arguments. Means and totals, for example, are asymptotically normally distributed by the Central Limit Theorem. @sarndal2003model and @lohr2009sampling provide thorough reviews of the design-based approach.

@@

The model-based approach assumes the data are a random realization of a data-generating process. Randomness is often incorporated through distributional assumptions on this process. Instead of estimating fixed but unknown parameters (as in the design-based approach), the goal of model-based inference in the spatial context is often \emph{prediction} of an unknown quantity. For example, suppose the realized mean of all population units is the quantity of interest. Instead of \emph{estimating} a fixed unknown mean, we are \emph{predicting} the value of the mean, a random variable. We know that if we sampled all population units, we would have an exact prediction for the mean of our one realized process, without any uncertainty. But we are often not interested in the true, unknown mean of the underlying process.

Assuming the data is a realization of a specific data-generating process yields predictors that are linked to distributional assumptions. These distributional assumptions are used to derive prediction intervals. The distributional assumptions allow the prediction intervals to be more precise. @cressie1993statistics and @schabenberger2017statistical provide reviews of model-based approaches for spatial data.

<!-- Figure 1a. Brus (2020): Data is fixed. In a finite population example, show a 3d surface that can be generated by anything. If we repeatedly sample the surface, then 95% of all 95% CIs will contain the true mean, which never changes. -->

```{r fig1, out.width = "100%", fig.cap = "A comparison of sampling under the design-based and model-based frameworks. Points circled are those that are sampled. In the top row, we have one fixed population, and three random samples. In the bottom row, we have three realizations of the same spatial process sampled at the same locations."}
set.seed(10102021)

@@

Description of Figure \ref{fig:fig1} goes here.

<!-- Figure 1b. Spatial process is fixed. In a finite population example, show 10 3d surfaces that are generated from some model. If we repeatedly generate the surface and obtain a sample, then 95% of all 95% PIs will contain the realized means. The realized mean changes from surface to surface and it's not necessarily the case that 95% of all 95% PIs will contain the true, underlying mean. -->

>>>>

Treating the data as fixed and incorporating randomness through the sampling design (top row of Figure \ref{fig:fig1} ((cite Brus 2021 here since our figure is similar?))) yields estimators having very few other assumptions. Confidence intervals for these types of estimators are typically derived using limiting arguments. Means and totals, for example, are asymptotically normally distributed by the Central Limit Theorem. If we repeatedly sample the surface, then 95% of all 95% confidence intervals constructed from a procedure with appropriate coverage will contain the true, fixed mean. @sarndal2003model and @lohr2009sampling provide thorough reviews of the design-based approach.

@@

The model-based approach assumes the data are a random realization of a data-generating process. Randomness is often incorporated through distributional assumptions on this process and need not be incorporated through random sampling (bottom row of Figure \ref{fig:fig1}). Instead of estimating fixed but unknown parameters (as in the design-based approach), the goal of model-based inference in the spatial context is often \emph{prediction} of an unknown quantity. For example, suppose the realized mean of all population units is the quantity of interest. Instead of \emph{estimating} a fixed unknown mean, we are \emph{predicting} the value of the mean, a random variable. We know that if we sampled all population units, we would have an exact prediction for the mean of our one realized process, without any uncertainty.

Assuming the data is a realization of a specific data-generating process yields predictors that are linked to distributional assumptions. These distributional assumptions are used to derive prediction intervals. The distributional assumptions allow the prediction intervals to be more precise. If we repeatedly generate the response values from a fixed spatial process and obtain a sample, then 95% of all 95% prediction intervals constructed from a procedure with appropriate coverage will contain their respective realized means. @cressie1993statistics and @schabenberger2017statistical provide reviews of model-based approaches for spatial data.

```{r fig1, out.width = "100%", fig.cap = "A comparison of sampling under the design-based and model-based frameworks. Points circled are those that are sampled. In the top row, we have one fixed population, and three random samples of n = 4. The response values at each site are fixed, but we obtain different estimates for the mean response because the randomly sampled sites vary from sample to sample. In the bottom row, we have three realizations of the same spatial process sampled at the same locations. The spatial process generating the response values has a single mean, but the realized mean is different in each of the three panels."}
set.seed(10102021)

@@

### 2021-10-10 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.tex
revised text for first conceptual figure.

Treating the data as fixed and incorporating randomness through the
sampling design yields estimators having very few other assumptions.
Confidence intervals for these types of estimators are typically derived
using limiting arguments. Means and totals, for example, are
asymptotically normally distributed by the Central Limit Theorem.
Särndal et al. (2003) and Lohr (2009) provide thorough reviews of the
design-based approach.

@@

The model-based approach assumes the data are a random realization of a
data-generating process. Randomness is often incorporated through
distributional assumptions on this process. Instead of estimating fixed
but unknown parameters (as in the design-based approach), the goal of
model-based inference in the spatial context is often \emph{prediction}
of an unknown quantity. For example, suppose the realized mean of all
population units is the quantity of interest. Instead of
\emph{estimating} a fixed unknown mean, we are \emph{predicting} the
value of the mean, a random variable. We know that if we sampled all
population units, we would have an exact prediction for the mean of our
one realized process, without any uncertainty. But we are often not
interested in the true, unknown mean of the underlying process.

Assuming the data is a realization of a specific data-generating process
yields predictors that are linked to distributional assumptions. These
distributional assumptions are used to derive prediction intervals. The
distributional assumptions allow the prediction intervals to be more
precise. Cressie (1993) and Schabenberger and Gotway (2017) provide
reviews of model-based approaches for spatial data.

\begin{figure}
\includegraphics[width=1\linewidth]{SpatialDVM_Manuscript_files/figure-latex/fig1-1} \caption{A comparison of sampling under the design-based and model-based frameworks. Points circled are those that are sampled. In the top row, we have one fixed population, and three random samples. In the bottom row, we have three realizations of the same spatial process sampled at the same locations.}\label{fig:fig1}
\end{figure}

Description of Figure \ref{fig:fig1} goes here.

>>>>

Treating the data as fixed and incorporating randomness through the
sampling design (top row of Figure \ref{fig:fig1} ((cite Brus 2021 here
since our figure is similar?))) yields estimators having very few other
assumptions. Confidence intervals for these types of estimators are
typically derived using limiting arguments. Means and totals, for
example, are asymptotically normally distributed by the Central Limit
Theorem. If we repeatedly sample the surface, then 95\% of all 95\%
confidence intervals constructed from a procedure with appropriate
coverage will contain the true, fixed mean. Särndal et al. (2003) and
Lohr (2009) provide thorough reviews of the design-based approach.

@@

The model-based approach assumes the data are a random realization of a
data-generating process. Randomness is often incorporated through
distributional assumptions on this process and need not be incorporated
through random sampling (bottom row of Figure \ref{fig:fig1}). Instead
of estimating fixed but unknown parameters (as in the design-based
approach), the goal of model-based inference in the spatial context is
often \emph{prediction} of an unknown quantity. For example, suppose the
realized mean of all population units is the quantity of interest.
Instead of \emph{estimating} a fixed unknown mean, we are
\emph{predicting} the value of the mean, a random variable. We know that
if we sampled all population units, we would have an exact prediction
for the mean of our one realized process, without any uncertainty.

Assuming the data is a realization of a specific data-generating process
yields predictors that are linked to distributional assumptions. These
distributional assumptions are used to derive prediction intervals. The
distributional assumptions allow the prediction intervals to be more
precise. If we repeatedly generate the response values from a fixed
spatial process and obtain a sample, then 95\% of all 95\% prediction
intervals constructed from a procedure with appropriate coverage will
contain their respective realized means. Cressie (1993) and
Schabenberger and Gotway (2017) provide reviews of model-based
approaches for spatial data.

\begin{figure}
\includegraphics[width=1\linewidth]{SpatialDVM_Manuscript_files/figure-latex/fig1-1} \caption{A comparison of sampling under the design-based and model-based frameworks. Points circled are those that are sampled. In the top row, we have one fixed population, and three random samples of n = 4. The response values at each site are fixed, but we obtain different estimates for the mean response because the randomly sampled sites vary from sample to sample. In the bottom row, we have three realizations of the same spatial process sampled at the same locations. The spatial process generating the response values has a single mean, but the realized mean is different in each of the three panels.}\label{fig:fig1}
\end{figure}
